\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-08 k-planar.tex
Quitting for the day

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs (and other more general classes). Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.

@@

\cref{modern-major-general,PartitionProduct,MakeWidth1} immediately imply \cref{ShortcutProduct} in the introduction.

@@

\cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with a weaker bound on the layered width. We prove the stronger bound in \cref{sec-k-planar}. Also note that \cref{k-planar,PartitionProduct} imply \cref{kPlanarBasic} with the $O(k^5)$ bound on the treewidth of $H$, as promised.

In the special case $k=1$ we obtain better constants and an additional property (planarity) of $H$ (see \cref{sec-1-planar} for the proof):

@@

We prove that $(g,k)$-planar graphs admits bounded layered partitions.

@@

\note{DW}{should this subsection be moved to the end of \cref{Tools}?}

@@

For convenience, it will be helpful to assume that $\mathcal{P}$ contains a length-1 $vw$-shortcut for every edge $vw\in E(G)$.  This assumption has no effect on the resulting graph $G^{\mathcal{P}}$ but avoids special cases in some of our proofs.

We begin with a fairly standard lemma about normalized tree decompositions. 
For a rooted tree $T$ and any node $x\in V(T)$, let  
$$T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}]$$ 
be the maximal subtree of $T$ rooted at $x$.

\begin{lem}\lemlabel{nice-decomposition}
  For any graph $H$ of treewidth $t$, there is a rooted tree $T$ with $V(T)=V(H)$ and a $T$-decomposition $(B_x:x\in V(T))$ of $H$ that has following additional properties:  
  \begin{compactenum}[(T1)]
    % \item\tlabel{rooted}\tlabel{first} $T$ is rooted at some node $x_0\in V(T)$ with $B_{x_0}=\emptyset$.
    % \item\tlabel{node-set} $V(T)= V(H)$.
    % \item\tlabel{diff} For each edge $xy\in E(T)$, $|B_x\ominus B_y|\le 1$.
    \item\tlabel{subtree-root} For each $x\in V(H)$, the subtree $T[x]:=T[\{y\in V(T):x\in B_y\}]$ is rooted at $x$.
    \item\tlabel{ancestor-edge}\tlabel{last} For each edge $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
  \end{compactenum}
\end{lem}

\begin{proof}
  Begin with any width-$t$ tree decomposition $(B_x:x\in V(T_0))$ of $H$ that uses some rooted tree $T_0$.  Select any node $x\in V(T_0)$, add a leaf $x_0$, with $B_{x_0}=\emptyset$, adjacent to $x$ and root $T_0$ at $x_0$.  Let $f:V(H)\to V(T)$ be function that maps each $x\in V(H)$ onto the root of the subtree $T_0[x]:=T_0[\{y\in V(T_0): x\in B_y]$.  If $f$ is not one-to-one, then select some distinct pair $x,y\in V(H)$ with $a:=f(x)=f(y)$.  Subdivide the edge between $a$ and its parent in $T$ by introducing a new node $a'$ with $B_{a'}=B_{a}\setminus\{x\}$.  This modification reduces the number of distinct pairs $x,y\in V(H)$ with $f(x)=f(y)$, so repeatedly performing this modification will eventually produce a tree-decomposition $(B_x:x\in V(T_0))$ of $H$ in which $f$ is one-to-one.

@@

For the remainder of this section, $G$ is a graph and $(\mathcal{L}:=\langle V_1,\ldots,V_h\rangle,\mathcal{Y}:=(Y_x: x\in V(H)))$ is a layered partition of $G$ of layered width $\ell$ where the treewidth of the quotient graph $H=G/\mathcal{Y}$ is at most $t$.  In addition, $(B_x:x\in V(T))$ is a tree-decomposition of $H$ obtained by applying \lemref{nice-decomposition} to $H$.  The following lemma shows how to interpret these objects as a hierarchical decomposition of $G$.

@@

As in the rest of this section, $(\mathcal{L},\mathcal{Y})$ denotes a layered $H$-partition of $G$; $T$ is a tree used in a tree-decomposition $(B_x:x\in V(T))$ of $H:=G/\mathcal{Y}$ satisfying the conditions of \lemref{nice-decomposition}; and, for each $x\in V(T)$, $Y_x$, $V_x$, $F_x$, and $N_x$ are defined as in \clmref{generalized-tripod}.

@@

For each $x\in V(T)$, we define the \emph{separator}
\[  
     S_x := \{v\in V(G): a(v)= x\} \enspace .
\]
Observe that $\mathcal{S}:=(S_x : x\in V(T))$ is a partition of $V(G)$.  We let $I:=G^\mathcal{P}/\mathcal{S}$ denote the resulting quotient graph and we let $V(I)\subseteq V(T)$ in the obvious way, so that each $x\in V(I)$ is the vertex obtained by contracting $S_x$ in $G^{\mathcal{P}}$. (Nodes $x\in V(T)$ with $S_x=\emptyset$ do not contribute a vertex to $I$.)

@@

\begin{proof}
  For the sake of contradiction, assume otherwise, so there exists some $v\in S_x\setminus V_x$.

  By \yref{y-subsets}, $Y_x\subseteq V_x$, so $v\not\in Y_x$.  Therefore, $\mathcal{P}$ contains a path $P$ that crosses $x$.  The path $P$ contains a subpath $v_0,v_1,\ldots,v_{r}$ such that $v=v_0$ and $v_r\in Y_x$. Since $v\not\in V_x$ and $v_r\in Y_x\subseteq V_x$, \yref{separator}, implies that $v_i\in F_x$ for some $i\in\{0,\ldots,r-1\}$. Now \yref{ancestor-edge} implies $v_i\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  Therefore, either $v\in Y_a$ or $P$ crosses $a$. But this implies that $a(v)$ is an ancestor of $a$, contradicting the assumption that $v\in S_x$.
\end{proof}

@@

\begin{clm}\clmlabel{general-width}
  For each $i\in\{1,\ldots,h\}$ and each $x\in V(I)$, $|S_x\cap V_i|\le d\ell(k^2+k+4)$.  
\end{clm}

\begin{proof}
  We will count the number of vertices in $S_x\cap V_i$ by upper-bounding the number of vertices contributed to $S_x\cap V_i$ by each vertex $w\in Y_x$.
  If $w\in Y_x\cap V_i$ and no path in $\mathcal{P}$ includes $w$ as an internal vertex then $w$ contributes only one vertex, itself, to $S_x\cap V_i$.

@@

  For any $j$, the number of vertices $w\in V_{i+j}\cap Y_x$ is at most $\ell$. Each such vertex $w$ is an internal vertex of at most $d$ paths in $\mathcal{P}$. Therefore, 
  \[  |S_x\cap V_i|\le d\ell\cdot\left(k+1 + k + \sum_{j=2}^k 2(k-j)\right) %= d(2k+1) + \sum_{i=1}^{k-2} i
      = d\ell(k^2 +k + 4) \enspace . \qedhere
  \]
\end{proof}

\begin{clm}\clmlabel{i-ancestor}
  For each edge $xy\in E(I)$, one of $x$ or $y$ is a $T$-ancestor of the other.
\end{clm}

\begin{proof}
  Suppose, for the sake of contradiction, that neither $x$ nor $y$ is a $T$-ancestor of the other.  Since $xy\in E(I)$, $G^\mathcal{P}$ contains an edge $vw$ with $v\in S_x$ and $w\in S_y$.  Since $xy\in E(G^{\mathcal{P}})$,  $\mathcal{P}$ contains a $vw$-shortcut $P$.  By \clmref{s-subset}, $v\in V_x$ and $w\in V_y$.  By \yref{containment-i}, if neither $x$ nor $y$ is a $T$-ancestor of the other, then $V_x$ and $V_y$ are disjoint.  By \yref{ancestor-edge} $N_x$ and $V_y$ are also disjoint.  By \yref{separator} $P$ contains an internal vertex $v'\in F_x$.  By \yref{ancestor-edge}, $v'\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  But this implies that $v\in S_{a'}$ for some $T$-ancestor $a'$ of $a$, contradicting the assumption that $v\in S_x$.
\end{proof}

\begin{clm}\clmlabel{general-bag-size}
  The graph $I$ has a tree-decomposition in which every bag has size at most $\binom{k+t}{t}$.
\end{clm}

\begin{proof}
  For the tree-decomposition $(C_x:x\in V(T))$ of $I$ we use the same tree $T$ used in the tree-decomposition $(B_x:x\in V(T))$ of $H$. For each node $x$ of $T$, $C_x$ contains $x$ as well as every $T$-ancestor $a$ of $x$ such that $I$ contains an edge $ax'$ where $x$ is a $T$-ancestor of $x'$ (including the possibility that $x=x'$).
  \clmref{i-ancestor} ensures that, for every edge $ax'\in E(I)$, $a,x'\in C_{x'}$ and the connectivity of $T[a]:=T[\{x\in V(T):a\in C_x]$ is obvious.  Therefore $(C_x:x\in V(T))$ is indeed a tree-decomposition of $I$.  What remains is to bound the size, $\max\{|C_x|:x\in V(T)\}$, of the largest bag.

@@

  If $x_\delta\in C_x$, it is because $x_\delta x'\in E(I)$ for some $T$-descendant $x'$ of $x$.  This implies $G^{\mathcal{P}}$ contains an edge $vw$ with $v\in S_{x'}$ and $w\in S_{x_\delta}=S_\delta$.  This implies that $\mathcal{P}$ a $vw$-shortcut $P_{vw}$.  Let $v'$ be the second-last vertex of $P_{vw}$ (so $v'w\in E(G)$).

@@

Let $G$ be a $k$-plane graph.  We will assume, for ease of exposition, that any point $p\in\R^2$ is involved in at most one crossing $(p,vw,xy)$ of $G$. This assumption is not critical since it can be enforced by a slight deformation of the edges of $G$ and the resulting (deformed) graph is also $k$-plane.

Let $G_0$ be the plane graph obtained by adding a dummy vertex at each crossing in $G$. In this way, each edge $vw\in E(G)$ corresponds naturally to a path $P_{vw}$ of length at most $k+1$ in $G_0$.  Let $\PP := \{P_{vw}: vw\in E(G)\}$. Observe that $\PP$ is a $(k+1,2)$-shortcut for $G_0$ and that $G_0^{\mathcal{P}}\supseteq G$.  Specifically, $G_0^{\mathcal{P}}$ contains every edge and vertex of $G$ as well as the dummy vertices in $V(G_0)\setminus V(G)$ and their incident edges.

Since $G_0$ is planar, the second part of \cref{PlanarPartition} implies that $G_0$ has an $H$-partition of layered width 3 for some planar graph $H$ of treewidth at most 3.  Applying \cref{modern-major-general} to $G_0$ and $\mathcal{P}$ immediately implies that $G$ (an arbitrary $k$-planar graph) has a layered $H$-partition of layered width $3(k^3+3k)$ for some graph $H$ of treewidth at most $\binom{k+4}{3}-1$.

We can reduce the layered width of the $H$-partition of $G$ from $O(k^3)$ to $O(k^2)$ by observing that the dummy vertices in $V(G_0)\setminus V(G)$ do not contribute to the layered width of this partition.  In this setting, the proof of \cref{clm:general-width} is simpler since each vertex $w\in Y_x$ contributes at most two vertices to $V_i\cap Y_x$.  More precisely, a path $P\in\mathcal{P}$ containing an internal (dummy) vertex $w\in Y_x\cap (V_{i-j}\cup V_{i+j})$
\begin{compactenum}
  \item Contributes at most two vertices to $S_x\cap V_i$ for $j\in\{0,\ldots,\floor{(k+1)/2}$;
  \item Contributes at most one vertex to $S_x\cap V_j$ for $j\in\{\floor{(k+1)/2},\ldots,k+1\}$; 
  \item Contributes no vertices to $S_x\cap V_j$ for $j > k+1$.
\end{compactenum}
Redoing the calculation at the end of the proof of \cref{clm:general-width} then yields 
\begin{align*}
    |S_x\cap Y_i| &\le d\ell\left( 
       2 
       + 4\left\lfloor\frac{k+1}{2}\right\rfloor 
       + 2\left(k+1-\left\lfloor\frac{k+1}{2}\right\rfloor\right)
       \right)\\
       & =
       d\ell\left(
         2k + 2\left\lfloor\frac{k+1}{2}\right\rfloor + 4
       \right) \\
       & \le
       d\ell(3k+5) = 18k+30 \enspace .
\end{align*}    
With this change, the layered width of the partition given by \cref{modern-major-general} becomes $18k^2+30k$.  This establishes \cref{k-planar}.

@@

\note{PM}{I removed (commented out) the $(g,1)$-specialization. It requires cutting surfaces.  For reference, the statement was:
\begin{thm}
\label{g-1-planar}
Every $(g,1)$-planar graph has a layered $H$-partition of layered width at most $\max\{48g, 30\}$ in which $H$ has treewidth at most $4$.  
\end{thm}
}

@@

@@

For an integer $k\geq 0$, a \textit{$k$-queue layout} of a graph $G$ consists of a linear ordering $\preceq$ of $V(G)$ and a partition $\{E_1,E_2,\dots,E_k\}$ of $E(G)$, such that for $i\in\{1,2,\dots,k\}$, no two edges in $E_i$ are nested with respect to $\preceq$. That is, it is not the case that $v\prec x \prec y \prec w$ for edges $vw,xy\in E_i$. The \textit{queue-number} of a graph $G$, denoted by $\qn(G)$, is the minimum integer $k$ such that $G$ has a $k$-queue layout. Queue-number was introduced by \citet{HLR92}, who famously conjectured that planar graphs have bounded queue-number. \citet{dujmovic.joret.ea:planar} recently solved this question using \cref{PlanarPartition} and the following lemma. Indeed, resolving this question was the motivation for the development of \cref{PlanarPartition}.

\begin{lem}[\citep{dujmovic.joret.ea:planar}]
\label{qn}
If a graph $G$ has a layered $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$, then $\qn(G)\le 3 \ell (2^t-1) + \floor{\tfrac{3}{2}\ell}$.
\end{lem}

\cref{qn,g-k-planar} imply that $(g,k)$-planar graphs have queue-number at most $g 2^{O(k^3)}$. 
Note that \citet{dujmovic.joret.ea:planar} previously proved the bound of 
$O(g^{k+2})$ using \cref{SurfacePartition} and an ad-hoc method. Our result provides a better bound when $g>2^{k^2}$. 
%In particular, in the case for $k=1$, \cref{qn,1-planar,g-1-planar} imply:
%
%\begin{prop}
%For $g\geq 1$, every $(g,1)$-planar graph has queue-number at most $3 \times 48g (2^4-1) + \floor{\tfrac{3}{2} \times 48g }= 2232 g $.
%\end{prop}
%
%The previous best bound on the queue-number of $(g,1)$-planar graphs was $2(40g+490)^{3}$ due to \citet{dujmovic.joret.ea:planar}. For $g=0$ we can improve further. \citet{ABGKP18} proved that every planar graph with treewidth at most $3$ has queue-number at most $5$. Thus the graph $H$ in \cref{1-planar} has queue-number at most $5$. \cref{qn,1-planar} then imply:
In the case of 1-planar graphs we can improve further. \citet{ABGKP18} proved that every planar graph with treewidth at most $3$ has queue-number at most $5$. Thus the graph $H$ in \cref{1-planar} has queue-number at most $5$. \cref{qn,1-planar} then imply:

\begin{prop}
Every 1-planar graph has queue-number at most $3 \times 30 * 5 + \floor{\tfrac{3}{2} \times 30} = 495$. \end{prop}

@@

\noindent\note{PM}{Double-check the calculation for \cref{k-planar} and \cref{g-k-planar} and then fix the constants here.}\note{DW}{done}

@@

The following lemma is due to \citet{micek:personal}. We include the proof  for completeness. \note{DW}{can we replace by a result of \citet{micek:personal}?}

>>>>

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs (and other more general classes). Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.\note{PM}{I must be missing something.  The planarized graph $G'$ has $O(kn)$ vertices.  Let $S$ be the set of non-dummy vertices and take a separation $(A,B)$ of $G'$ with $|A\cap B|=O(\sqrt{kn})$, $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$.  Now take $X=A\cap B$ plus the endpoints of every edge of $G$ that contains a dummy vertex in $A\cap B$.  $|X|\le 5|A\cap B|=O(\sqrt{kn})$ and $X$ is a separator of $G$.}

@@

\cref{modern-major-general,MakeWidth1} immediately imply \cref{ShortcutProduct} in the introduction.

@@

\cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with the weaker bound of $(2\times 3)((k+1)^3+3(k+1))=6k^3 + 18k^2 + 36k+24$ on the layered width. We prove the stronger bound in \cref{sec-k-planar}. 
%Also note that \cref{k-planar,MakeWidth1} imply \cref{kPlanarBasic} with the $O(k^5)$ bound on the treewidth of $H$, as promised. 
% PM: I removed the last line because it's already stated in the line preceding the theorem

In the important special case of $k=1$ we obtain better constants and an additional property (planarity) of $H$ (see \cref{sec-1-planar} for the proof):

@@

We prove that $(g,k)$-planar graphs admit bounded layered partitions.

@@

\note{DW}{should this subsection be moved to the end of \cref{Tools}?}\note{PM}{No, I don't think so. The information given in Section 2 isn't detailed enough to compare with layered treewidth versus layered partitions.}

@@

For convenience, it will be helpful to assume that $\mathcal{P}$ contains a length-1 $vw$-shortcut for every edge $vw\in E(G)$.  Since $G^\PP$ is defined to be a supergraph of $G$, this assumption has no effect on $G^{\mathcal{P}}$ but avoids special cases in some of our proofs.

For a rooted tree $T$ rooted at some node $x_0\in V(T)$, we we say that a node $a\in V(T)$ is a \emph{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \emph{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in v(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \emph{strict} $T$-ancestor of $x$ if $a\neq x$. For a rooted tree $T$ and any node $x\in V(T)$, let  
\[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
be the maximal subtree of $T$ rooted at $x$.

We begin with a fairly standard lemma about normalized tree decompositions.

\begin{lem}\lemlabel{nice-decomposition}
  For any graph $H$ of treewidth $t$, there is a rooted tree $T$ with $V(T)=V(H)$ and a $T$-decomposition $(B_x:x\in V(T))$ of $H$ that has following additional properties:  
  \begin{compactenum}[(T1)]
    % \item\tlabel{rooted}\tlabel{first} $T$ is rooted at some node $x_0\in V(T)$ with $B_{x_0}=\emptyset$.
    % \item\tlabel{node-set} $V(T)= V(H)$.
    % \item\tlabel{diff} For each edge $xy\in E(T)$, $|B_x\ominus B_y|\le 1$.
    \item\tlabel{subtree-root} For each node $x\in V(H)$, the subtree $T[x]:=T[\{y\in V(T):x\in B_y\}]$ is rooted at $x$.
    \item\tlabel{ancestor-edge}\tlabel{last} For each edge $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
  \end{compactenum}
\end{lem}

\begin{proof}
  Begin with any width-$t$ tree decomposition $(B_x:x\in V(T_0))$ of $H$ that uses some tree $T_0$.  Select any node $x\in V(T_0)$, add a leaf $x_0$, with $B_{x_0}=\emptyset$, adjacent to $x$ and root $T_0$ at $x_0$.  Let $f:V(H)\to V(T)$ be the function that maps each $x\in V(H)$ onto the root of the subtree $T_0[x]:=T_0[\{y\in V(T_0): x\in B_y]$.  If $f$ is not one-to-one, then select some distinct pair $x,y\in V(H)$ with $a:=f(x)=f(y)$.  Subdivide the edge between $a$ and its parent in $T$ by introducing a new node $a'$ with $B_{a'}=B_{a}\setminus\{x\}$.  This modification reduces the number of distinct pairs $x,y\in V(H)$ with $f(x)=f(y)$, so repeatedly performing this modification will eventually produce a tree-decomposition $(B_x:x\in V(T_0))$ of $H$ in which $f$ is one-to-one.

@@

For the remainder of this section, $G$ is a graph and $(\mathcal{L}:=\langle V_1,\ldots,V_h\rangle,\mathcal{Y}:=(Y_x: x\in V(H)))$ is a layered partition of $G$ of layered width $\ell$ where the treewidth of the quotient graph $H=G/\mathcal{Y}$ is at most $t$.  In addition, $\mathcal{T}:=(B_x:x\in V(T))$ is a tree-decomposition of $H$ obtained by applying \lemref{nice-decomposition} to $H$.  The following lemma shows how to interpret the pair $(\mathcal{Y},\mathcal{T})$ as a hierarchical decomposition of $G$.  Refer to \cref{fig:generalized-tripod}

\begin{figure}
  \begin{center}
    \includegraphics{figs/tripoddo}
  \end{center}
  \caption{The sets $Y_x$, $F_x$, and $N_x$ associated with $x\in V(T)$
  and the ancestors $a_1,\ldots,a_{t'}$ of $X$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.}
  \label{fig:generalized-tripod}
\end{figure}

@@

In the rest of this section, 
%$(\mathcal{L},\mathcal{Y})$ denotes a layered $H$-partition of $G$;
$\mathcal{T}=(B_x:x\in V(T))$ is a tree-decomposition of $H:=G/\mathcal{Y}$ satisfying the conditions of \lemref{nice-decomposition} and, for each $x\in V(T)$, $Y_x$, $V_x$, $F_x$, and $N_x$ are defined as in \clmref{generalized-tripod}.

@@

For each $x\in V(T)$, we define the \emph{separator}
\[  
     S_x := \{v\in V(G): a(v)= x\} \enspace .
\]
Observe that $\mathcal{S}:=(S_x : x\in V(T))$ is a partition of $V(G)$.  We let $J:=G^\mathcal{P}/\mathcal{S}$ denote the resulting quotient graph and we let $V(J)\subseteq V(T)$ in the obvious way, so that each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\mathcal{P}}$. (Nodes $x\in V(T)$ with $S_x=\emptyset$ do not contribute a vertex to $J$.)

@@

\begin{proof}
  For the sake of contradiction, assume otherwise, so there exists some $v\in S_x\setminus V_x$. By \yref{y-subsets}, $Y_x\subseteq V_x$, so $v\not\in Y_x$.  Therefore, $\mathcal{P}$ contains a path $P$ that crosses $x$.  The path $P$ contains a subpath $v_0,v_1,\ldots,v_{r}$ such that $v=v_0$ and $v_r\in Y_x$. Since $v\not\in V_x$ and $v_r\in Y_x\subseteq V_x$, \yref{separator}, implies that $v_i\in F_x$ for some $i\in\{0,\ldots,r-1\}$. Now \yref{ancestor-edge} implies $v_i\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  Therefore, either $v\in Y_a$ or $P$ crosses $a$. But this implies that $a(v)$ is a $T$-ancestor of $a$, which is a strict $T$-ancestor of $x$, contradicting the assumption that $v\in S_x$.
\end{proof}

@@

\begin{clm}\clmlabel{general-width}
  For each $i\in\{1,\ldots,h\}$ and each $x\in V(J)$, $|S_x\cap V_i|\le d\ell(k^2+3)$.  
\end{clm}

\begin{proof}
  We will count the number of vertices in $S_x\cap V_i$ by upper-bounding the number of vertices contributed to $S_x\cap V_i$ by each vertex $w\in Y_x$. Refer to \cref{contribute}.
  If $w\in Y_x\cap V_i$ and no path in $\mathcal{P}$ includes $w$ as an internal vertex then $w$ contributes only one vertex, itself, to $S_x\cap V_i$.
  \begin{figure}
    \begin{center}
      \includegraphics{figs/contribute}
    \end{center}
    \caption{A path $P$ containing an internal vertex $w\in Y_x\cap V_{i-j}$.}
    \label{contribute}
  \end{figure}

@@

  For any $j$, the number of vertices $w\in V_{i+j}\cap Y_x$ is at most $\ell$. Each such vertex $w$ is an internal vertex of at most $d$ paths in $\mathcal{P}$. Therefore, 
  \[  |S_x\cap V_i|\le d\ell\cdot\left(k+1 + 2k + \sum_{j=2}^k 2(k-j)\right) %= d(2k+1) + \sum_{i=1}^{k-2} i
      = d\ell(k^2 +3) \enspace . \qedhere
  \]
\end{proof}

\begin{clm}\clmlabel{i-ancestor}
  For each edge $xy\in E(J)$, one of $x$ or $y$ is a $T$-ancestor of the other.
\end{clm}

\begin{proof}
  Suppose, for the sake of contradiction, that neither $x$ nor $y$ is a $T$-ancestor of the other.  Since $xy\in E(J)$, $G^\mathcal{P}$ contains an edge $vw$ with $v\in S_x$ and $w\in S_y$.  Since $xy\in E(G^{\mathcal{P}})$,  $\mathcal{P}$ contains a $vw$-shortcut $P$.  By \clmref{s-subset}, $v\in V_x$ and $w\in V_y$.  By \yref{containment-i}, if neither $x$ nor $y$ is a $T$-ancestor of the other, then $V_x$ and $V_y$ are disjoint.  By \yref{ancestor-edge} $N_x$ and $V_y$ are also disjoint.  By \yref{separator} $P$ contains an internal vertex $v'\in F_x$.  By \yref{ancestor-edge}, $v'\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  But this implies that $v\in S_{a'}$ for some $T$-ancestor $a'$ of $a$, contradicting the assumption that $v\in S_x$.
\end{proof}

\begin{clm}\clmlabel{general-bag-size}
  The graph $J$ has a tree-decomposition in which every bag has size at most $\binom{k+t}{t}$.
\end{clm}

\begin{proof}
  For the tree-decomposition $(C_x:x\in V(T))$ of $J$ we use the same tree $T$ used in the tree-decomposition $(B_x:x\in V(T))$ of $H$. For each node $x$ of $T$, $C_x$ contains $x$ as well as every $T$-ancestor $a$ of $x$ such that $J$ contains an edge $ax'$ where $x$ is a $T$-ancestor of $x'$ (including the possibility that $x=x'$).
  \clmref{i-ancestor} ensures that, for every edge $ax'\in E(J)$, $a,x'\in C_{x'}$ and the connectivity of $T[a]:=T[\{x\in V(T):a\in C_x]$ is obvious.  Therefore $(C_x:x\in V(T))$ is indeed a tree-decomposition of $J$.  What remains is to bound the size, $\max\{|C_x|:x\in V(T)\}$, of the largest bag.

@@

  If $x_\delta\in C_x$, it is because $x_\delta x'\in E(J)$ for some $T$-descendant $x'$ of $x$.  This implies $G^{\mathcal{P}}$ contains an edge $vw$ with $v\in S_{x'}$ and $w\in S_{x_\delta}=S_\delta$.  This implies that $\mathcal{P}$ a $vw$-shortcut $P_{vw}$.  Let $v'$ be the second-last vertex of $P_{vw}$ (so $v'w\in E(G)$).

@@

In this section we show that the constants in \cref{modern-major-general} can be improved for the important special case of $1$-planar graphs. Let $G$ be a $k$-plane graph.  We will assume, for ease of exposition, that any point $p\in\R^2$ is involved in at most one crossing $(p,vw,xy)$ of $G$. This assumption is not critical since it can be enforced by a slight deformation of the edges of $G$ and the resulting (deformed) graph is also $k$-plane.

Let $G_0$ be the plane graph obtained by adding a dummy vertex at each crossing in $G$. In this way, each edge $vw\in E(G)$ corresponds naturally to a path $P_{vw}$ of length at most $k+1$ in $G_0$.  Let $\PP := \{P_{vw}: vw\in E(G)\}$. Observe that $\PP$ is a $(k+1,2)$-shortcut system for $G_0$ and that $G_0^{\mathcal{P}}\supseteq G$.  Specifically, $G_0^{\mathcal{P}}$ contains every edge and vertex of $G$ as well as the dummy vertices in $V(G_0)\setminus V(G)$ and their incident edges.

Since $G_0$ is planar, the second part of \cref{PlanarPartition} implies that $G_0$ has an $H$-partition of layered width 3 for some planar graph $H$ of treewidth at most 3.  Applying \cref{modern-major-general} to $G_0$ and $\mathcal{P}$ immediately implies that $G$ (an arbitrary $k$-planar graph) has a layered $H$-partition of layered width $6((k+1)^3+3(k+1))$ for some graph $H$ of treewidth at most $\binom{k+4}{3}-1$.

We can reduce the layered width of the $H$-partition of $G$ from $O(k^3)$ to $O(k^2)$ by observing that the dummy vertices in $V(G_0)\setminus V(G)$ do not contribute to the layered width of this partition.  In this setting, the proof of \cref{clm:general-width} is simpler since each vertex $w\in Y_x$ contributes at most two vertices to $V_i\cap Y_x$.  More precisely, a path $P\in\mathcal{P}$ containing an internal (dummy) vertex $w\in Y_x\cap (V_{i-j}\cup V_{i+j})$
\begin{compactenum}
  \item Contributes at most two vertices to $S_x\cap V_i$ for $j\in\{0,\ldots,\floor{(k+1)/2}\}$;
  \item Contributes at most one vertex to $S_x\cap V_j$ for $j\in\{\floor{(k+1)/2},\ldots,k+1\}$; 
  \item Contributes no vertices to $S_x\cap V_j$ for $j > k+1$.
\end{compactenum}
Redoing the calculation at the end of the proof of \cref{clm:general-width} then yields 
\begin{align*}
    |S_x\cap Y_i| &\le d\ell\left( 
       2 
       + 4\left\lfloor\frac{k+1}{2}\right\rfloor 
       + 2\left(k+1-\left\lfloor\frac{k+1}{2}\right\rfloor\right)
       \right)\\
       & =
       d\ell\left(
         2k + 2\left\lfloor\frac{k+1}{2}\right\rfloor + 4
       \right) \\
       & \le
       d\ell(3k+5) = 18k+30 \enspace .
\end{align*}    
With this change, the layered width of the partition given by \cref{modern-major-general} becomes $18k^2+30k$.  This establishes \cref{k-planar}.

@@

\note{PM}{I removed (commented out) the $(g,1)$-specialization. It requires cutting surfaces.  For reference, the statement was:
\begin{thm}
\label{g-1-planar}
Every $(g,1)$-planar graph has a layered $H$-partition of layered width at most $\max\{48g, 30\}$ in which $H$ has treewidth at most $4$.  
\end{thm}
}

@@

\note{PM}{I propose we leave it out of the SODA submission, but I'm happy to include it in the DCG/ArXiV submission (which we can hopefully finish in the next few days.)}

@@

For an integer $k\geq 0$, a \textit{$k$-queue layout} of a graph $G$ consists of a linear ordering $\preceq$ of $V(G)$ and a partition $\{E_1,E_2,\dots,E_k\}$ of $E(G)$, such that for $i\in\{1,2,\dots,k\}$, no two edges in $E_i$ are nested with respect to $\preceq$. That is, it is not the case that $v\prec x \prec y \prec w$ for edges $vw,xy\in E_i$. The \textit{queue-number} of a graph $G$, denoted by $\qn(G)$, is the minimum integer $k$ such that $G$ has a $k$-queue layout. Queue-number was introduced by \citet{HLR92}, who famously conjectured that planar graphs have bounded queue-number. \citet{dujmovic.joret.ea:planar} recently proved this conjecture using \cref{PlanarPartition} and the following lemma. Indeed, resolving this question was the motivation for the development of \cref{PlanarPartition}.

\begin{lem}[\citep{dujmovic.joret.ea:planar}]
\label{qn}
If a graph $G$ has a layered $H$-partition of layered width $\ell$ in which $H$ has queue number at most $q$, then $\qn(G)\le 3 \ell q + \floor{\tfrac{3}{2}\ell}$.
\end{lem}

Several authors have shown that queue-number is bounded by a function of treewidth. The best current result of this type is due to \cite{wiechert:on}

\begin{thm}\label{qn-tw}
    For every graph $G$, $\qn(G)\le 2^{\tw(G)}-1$.
\end{thm}

\cref{qn,qn-tw,g-k-planar} imply that $(g,k)$-planar graphs have queue-number at most $g 2^{O(k^3)}$. 
Note that \citet{dujmovic.joret.ea:planar} previously proved the bound of 
$O(g^{k+2})$ using \cref{SurfacePartition} and an ad-hoc method. Our result provides a better bound when $g>2^{k^2}$. 
%In particular, in the case for $k=1$, \cref{qn,1-planar,g-1-planar} imply:
%
%\begin{prop}
%For $g\geq 1$, every $(g,1)$-planar graph has queue-number at most $3 \times 48g (2^4-1) + \floor{\tfrac{3}{2} \times 48g }= 2232 g $.
%\end{prop}
%
%The previous best bound on the queue-number of $(g,1)$-planar graphs was $2(40g+490)^{3}$ due to \citet{dujmovic.joret.ea:planar}. For $g=0$ we can improve further. \citet{ABGKP18} proved that every planar graph with treewidth at most $3$ has queue-number at most $5$. Thus the graph $H$ in \cref{1-planar} has queue-number at most $5$. \cref{qn,1-planar} then imply:
In the case of 1-planar graphs we can improve further. \citet{ABGKP18} proved that every planar graph with treewidth at most $3$ has queue-number at most $5$. Thus the graph $H$ in \cref{1-planar} has queue-number at most $5$. \cref{qn,1-planar} then imply:

\begin{prop}
Every 1-planar graph has queue-number at most $3 \times 30 \times 5 + \floor{\tfrac{3}{2} \times 30} = 495$. \end{prop}

@@

% \noindent\note{PM}{Double-check the calculation for \cref{k-planar} and \cref{g-k-planar} and then fix the constants here.}\note{DW}{done}

@@

The following lemma is due to \citet{micek:personal}. We include the proof  for completeness. 
% \note{DW}{can we replace by a result of \citet{micek:personal}?}
% \note{PM}{I didn't see a statement of this exact result in the last draft Piotr sent.  Though their Theorem~7 is this proof for the special case of planar graphs.}
